Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2015-02-20 papers/wdmerger_I/paper.tex
Add uniform cube/sphere convergence test plot to paper.

Since the results of a merger simulation depend strongly on gravity,
it is important to check whether proper numerical convergence is
achieved for the Poisson solver. To do so, we created a simple test
that inserts a sphere of radius $R$ and uniform mass density $\rho$
onto our grid, and used CASTRO to calculate the gravitational
potential $\Phi$ of this setup. We ensure that $R$ is an integer
multiple of the grid spacing, and the center of the sphere is at the
origin. The problem domain for our simulations is $[-1.6, 1.6]^3$, and
we take $R = 0.5$ and $\rho = 10^3$. The zones with $r > R$ are filled
with an ambient material of very low density. We run this problem at
multiple resolutions corresponding to jumps by a factor of two. For
comparison, at each grid point we evaluate the analytical potential of
a uniform sphere, which can be easily determined using Gauss' law:
\begin{equation}
  \Phi_{\text{sphere}}(r) = -\frac{GM}{r} \times \begin{cases} (3R^2 - r^2)/(2 r^2) & r \leq R \\ 1 & r > R \end{cases},\label{eq:sphere-analytical}
\end{equation}
where $M = 4\pi R^3 / 3$ is the mass of the sphere. We measure the 
numerical error by calculating the $L^2$ norm of the error and 
normalizing it by the $L^2$ norm of the analytical solution:
\begin{equation}
  \text{Error} = \frac{\|\Phi - \Phi_{\text{sphere}}\|_2}{\|\Phi_{\text{sphere}}\|_2}.
\end{equation}

\begin{figure}
  \label{fig:gravity_convergence}
\end{figure}

The results of this test are plotted in Figure
\ref{fig:gravity_convergence}. We find that convergence is actually
substantially better than second-order, and that the difference is
largest at low resolutions. The explanation for this is that we are
attempting to model a spherical object on a rectangular grid. At very
low resolution, the object does not look very spherical. As the
resolution is increased, the total amount of mass on the grid, as well
as its location, will change as the sphere fills out. This means we
are combining the true accuracy bonus from increased resolution with
the artificial accuracy bonus from getting closer to solving the
problem we are supposed to be solving.

We can solve one of these two sources of error by evaluating Equation
\ref{eq:sphere-analytical} with a mass $M$ that corresponds to the
amount of mass actually on the grid, defined as the sum of the density
multiplied by the volume of each cell for all cells that have greater
than ambient density. The resolution study for this case is also
plotted in Figure \ref{fig:gravity_convergence}. We still obtain
convergence better than second-order, indicating that we still have
the geometrical problem.

The only way to fully eliminate this effect is to test a problem that
does not change with resolution. The obvious companion is a cube of
uniform density $\rho$, where now $R$ is half of the side length of
the cube. At each resolution we use the same $R$ as for the sphere,
which ensures that the cube always fills exactly the same fraction of
the domain and thus has the same mass, so the only improvement comes
from better sampling at higher resolution. The potential for this
object has been worked out analytically by \citet{waldvogel:1976} (see
also a similar result in \citealt{hummer:1996}). The potential is in
Equation 15 of that paper, though the last term is missing a factor of
$1/2$, which destroys the symmetry. Inserting this missing factor and
performing a simple coordinate transformation so that the center of
the cube is at the origin, the potential is
\begin{align}
  \Phi_{\text{cube}}(x,y,z) &= -G\rho\sum_{i,j,k=0}^1\left[x_i y_j\, \text{tanh}^{-1}\left(\frac{z_k}{r_{ijk}}\right)\right. \notag \\
  &+ \left. y_j z_k\, \text{tanh}^{-1}\left(\frac{x_i}{r_{ijk}}\right) + z_k x_i\, \text{tanh}^{-1}\left(\frac{y_j}{r_{ijk}}\right) \right.\notag \\
  &\left. - \frac{x_i^2}{2}\,\text{tan}^{-1}\left(\frac{y_j z_k}{x_i r_{ijk}}\right) - \frac{y_j^2}{2}\,\text{tan}^{-1}\left(\frac{z_k x_i}{y_j r_{ijk}}\right) \right. \notag \\
  &- \left. \frac{z_k^2}{2}\,\text{tan}^{-1}\left(\frac{x_i y_j}{z_k r_{ijk}}\right)\right]
\end{align}
where $x_0 = R + x$, $x_1 = R - x$, $y_0 = R + y$, $y_1 = R-y$, $z_0 =
R+z$, $z_1 = R-z$, and $r_{ijk} = \sqrt{x_i^2 + y_j^2 + z_k^2}$. In
\texttt{FORTRAN} and \texttt{C}, the inverse hyperbolic tangent is
\texttt{atanh} and the inverse tangent is \texttt{atan} (\textit{not}
\texttt{atan2}). This formula is valid both inside and outside the
cube. The normalized $L^2$ error norm for this problem is also shown
in Figure \ref{fig:gravity_convergence}, and indeed we see
near-perfect second-order scaling.

The main lesson here is that in a convergence study, it is important
to ensure that the physical problem does not change with
resolution. Since in the case of spherical objects on rectangular
grids the effect is to artificially boost convergence with resolution,
in a simulation with spherical objects like stars one can envision a
scenario of being fooled into believing apparently good convergence
results that are simply a convolution of artificially high
gravitational convergence and poor convergence in the hydrodynamics. A
convergence study in this case is only fully valid if there is reason
to be confident that this effect is negligible compared to other
factors.

Another factor that may affect convergence testing is easily seen in
Figure \ref{fig:gravity_convergence}, where at the highest resolution
the error flattens out. This occurs because of a limitation in the
accuracy of the multigrid solver. For problems of this size we find
empirically that the multigrid solver does not converge for error
tolerances smaller than about $10^{-11}$, which enforces a lower limit
on the total error. This is also something to be aware of in assessing
convergence for our full simulations.

>>>>

Since the results of a merger simulation depend strongly on gravity,
it is important to check whether proper numerical convergence is
achieved for the Poisson solver. To do so, we created a simple test
that inserts a sphere of radius $R$ and uniform mass density $\rho$
onto our grid, and used CASTRO to calculate the gravitational
potential $\Phi$ of this setup. We ensure that $R$ is an integer
multiple of the grid spacing, and the center of the sphere is at the
origin. The problem domain for our simulations is $[-1.6, 1.6]^3$, and
we take $R = 1.0 \text{cm}$ and $\rho = 10^3\ \text{g cm}^{-3}$. 
The zones with $r > R$ are filled with an ambient material of very low density 
$\rho = 10^{-8}\ \text{g cm}^{-3}$. We run this problem at multiple 
resolutions corresponding to jumps by a factor of two. For
comparison, at each grid point we evaluate the analytical potential of
a uniform sphere, which can be easily determined using Gauss' law:
\begin{equation}
  \Phi_{\text{sphere}}(r) = -\frac{GM}{r} \times \begin{cases} (3R^2 - r^2)/(2 r^2) & r \leq R \\ 1 & r > R \end{cases},\label{eq:sphere-analytical}
\end{equation}
where $M = 4\pi R^3 / 3$ is the mass of the sphere. We measure the 
numerical error by calculating the $L^2$ norm of the error and 
normalizing it by the $L^2$ norm of the analytical solution:
\begin{equation}
  \text{Error} = \frac{\|\Phi - \Phi_{\text{sphere}}\|_2}{\|\Phi_{\text{sphere}}\|_2}.
\end{equation}
We define the order of convergence $p$ between two simulations with a jump 
in resolution of integer factor $m > 1$ as
\begin{equation}
  p = \text{log}_{m}\left(\frac{\text{Error}_{\text{low}}}{\text{Error}_{\text{high}}}\right).
\end{equation}
Here $\text{Error}_{\text{low}}$ is the $L^2$ error at the lower resolution 
and $\text{Error}_{\text{high}}$ is the $L^2$ error at the higher resolution.
We expect the error to converge at $p = 2$ given the discretization we choose. 
For all simulations in this section and for all our main science simulations,
we choose a relative error tolerance of $10^{-10}$ to be satisfied in the multigrid solve.

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{plots/phi_comparison}
  \caption{Comparison of the CASTRO gravitational potential to the analytical solution for: 
    a sphere of uniform density; the same sphere, but with the potential normalized using the 
    actual amount of mass on the grid instead of the mass of a perfect sphere; and, a 
    cube of uniform density. Plotted also is a notional curve whose slope represents
    perfect second order convergence.\label{fig:gravity_convergence}}
\end{figure}

The results of this test are plotted in Figure
\ref{fig:gravity_convergence}. We find that at low resolution 
convergence is actually substantially better than second-order. The 
explanation for this is that we are attempting to model a spherical 
object on a rectangular grid. At very low resolution, the object does 
not look very spherical. As the resolution is increased, the total 
amount of mass on the grid, as well as its location, will change 
as the sphere fills out. This means we are combining the true accuracy 
bonus from increased resolution with the artificial accuracy bonus 
from getting closer to solving the problem we are supposed to be solving. 
On the other hand, at the highest resolution shown the error slightly increases.

We can solve one of these two sources of error by evaluating Equation
\ref{eq:sphere-analytical} with a mass $M$ that corresponds to the
amount of mass actually on the grid, defined as the sum of the density
multiplied by the volume of each cell for all cells whose density is 
greater than ambient. The resolution study for this case is also
plotted in Figure \ref{fig:gravity_convergence}. We still obtain
convergence slightly better than second-order, indicating that we 
still have the geometrical problem of the mass distribution changing.

The only way to fully eliminate this effect is to use a test problem that
does not change with resolution. The obvious companion is a cube of
uniform density $\rho$, where now $R$ is half of the side length of
the cube. At each resolution we use the same $R$ as for the sphere,
which ensures that the cube always fills exactly the same fraction of
the domain and thus has the same mass, so the only improvement comes
from better sampling at higher resolution. The gravitational potential for this
object has been worked out analytically by \citet{waldvogel:1976} (see
also a similar result in \citealt{hummer:1996}). The potential is in
Equation 15 of that paper, though the last term is missing a factor of
$1/2$, which destroys the symmetry. Inserting this missing factor and
performing a simple coordinate transformation so that the center of
the cube is at the origin, the potential is
\begin{align}
  \Phi_{\text{cube}}(x,y,z) &= -G\rho\sum_{i,j,k=0}^1\left[x_i y_j\, \text{tanh}^{-1}\left(\frac{z_k}{r_{ijk}}\right)\right. \notag \\
  &+ \left. y_j z_k\, \text{tanh}^{-1}\left(\frac{x_i}{r_{ijk}}\right) + z_k x_i\, \text{tanh}^{-1}\left(\frac{y_j}{r_{ijk}}\right) \right.\notag \\
  &\left. - \frac{x_i^2}{2}\,\text{tan}^{-1}\left(\frac{y_j z_k}{x_i r_{ijk}}\right) - \frac{y_j^2}{2}\,\text{tan}^{-1}\left(\frac{z_k x_i}{y_j r_{ijk}}\right) \right. \notag \\
  &- \left. \frac{z_k^2}{2}\,\text{tan}^{-1}\left(\frac{x_i y_j}{z_k r_{ijk}}\right)\right]
\end{align}
where $x_0 = R + x$, $x_1 = R - x$, $y_0 = R + y$, $y_1 = R-y$, $z_0 =
R+z$, $z_1 = R-z$, and $r_{ijk} = \sqrt{x_i^2 + y_j^2 + z_k^2}$. In
\texttt{Fortran} and \texttt{C}, the inverse hyperbolic tangent is
\texttt{atanh} and the inverse tangent is \texttt{atan} (\textit{not}
\texttt{atan2}). This formula is valid both inside and outside the
cube. The normalized $L^2$ error for this problem is also shown
in Figure \ref{fig:gravity_convergence}, and only for this problem 
do we obtain perfect second-order scaling at all resolutions.

The main lesson here is that in a convergence study, it is important
to ensure that the physical problem does not change with
resolution. Since in the case of spherical objects on rectangular
grids the effect may be to artificially boost convergence with resolution,
in a simulation with spherical objects like stars one can envision a
scenario of being fooled into believing apparently good convergence
results that are simply a convolution of artificially high
gravitational convergence and poor convergence in the hydrodynamics. A
convergence study in this case is only fully valid if there is reason
to be confident that this effect is negligible compared to other
factors.
